Compositional Certified Resource Bounds

### 2014-07-03 paper/main.tex
changes in section 2

\section{Informal Account}
\label{sec:inform}

In this section, we informally introduce the quantitative program
logic and the automatic amortized analysis for for Clight programs.
The purpose of this section is to build some intuition that should
make the following sections, which contain more details, easy to read.

@@

The idea that drives the design of our system is amortized
analysis~\cite{Tarjan-amort}.  Assume that a program $S$
executes on a starting state $\state_0$ and consumes $n$ resource units
of some quantity we are interested in.  We denote that by writing $(S,
\state_0) \Downarrow_n$.  The basic idea of amortized analysis is to
define a potential function $\Phi$ that maps program states to
non-negative numbers and to show that $\Phi(\state) \geq n$ if
$\state$ is a program state such that such that $(S, \state)
\Downarrow_n$.

Since programs can be composed we must enrich the evaluation judgement
$(S, \state) \Downarrow_{n}$ to talk about the state resulting from
the program execution.  We write $(S, \state) \Downarrow_n \state'$ to
say that $\state'$ is the program state after the execution of $S$ in
starting state $\sigma$.  We now use two potential functions, one that
applies before the execution, and one that applies after.  The two
functions must respect the relation $\Phi(\state) \ge n +
\Phi'(\state')$ for all states $\state$ and $\state'$ such that $(S,
\state) \Downarrow_n \state'$.  Intuitively, $\Phi(\state)$ must
provide enough \emph{potential} for both, paying for resource cost of
the computation and paying for the potential $\Phi'(\state')$ on the
resulting state $\state'$. That way, if $(\state, S_1) \Downarrow_n
\state'$ and $(\state', S_2) \Downarrow_m \state''$, we get
$\Phi(\state) \ge n + \Phi'(\state')$ and $\Phi'(\state') \ge m +
\Phi''(\state'')$.  This can be composed to $\Phi(\state) \ge (n + m)
+ \Phi''(\state'')$.  Note that the initial potential function $\Phi$
provides an upper bound on the resource consumption of the whole
program.  What we have observed is that, if we define $\htriple{\Phi}
{S}{\Phi'}$ to mean
$$
\forall \state\, n\, \state'.\, (\state, p) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state')
$$
then we get the following familiar looking rule.
$$
\RuleNolabel
{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
{\htriple{\Phi} {S_1; S_2}  {\Phi''}}
$$
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect systems
rules.  These rules enable interactive reasoning about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

We have implemented and proved sound such a quantitative Hoare for
CompCert Clight in the Coq Proof Assistant.  Because we use a shallow
embedding in Coq, every function that can be expressed in Coq can in
principle be used as a resource bound.  We can use the logic to prove
a concrete bound for a given resource metric, or a bound that is
parametric in a set of resource metrics.  It is also possible to
incorporate boolean preconditions into the bounds to express that the
bound is only valid for a certain class of inputs.  To this end, we
allow the potential function $\Phi$ in the pre and post conditions to
take non-negative number or $\infty$ (infinity) as values.  Infinity,
plays the same role as $\mathit{false}$ in Hoare logic.  Boolean assertions can
be embedded into potential functions by mapping $\mathit{false}$ to $\infty$ and
$\mathit{true}$ to $0$.

@@

\caption{Example derivation using amortized reasoning.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
  tick metric that assigns cost $n$ to \lstinline{tick(n)} statements and
  0 to other operations.
  }
\label{fig:xmplinc}
\end{figure}

It is for instance possible to derive the following quantitative
triple for a binary-search that holds for all stack metrics $M$.  A
stack metric assigns cost $0$ to all evaluation steps except function
calls.  Before a call $f(x)$, we consume $M_f > 0$ resources and
after the call we return $M_f$ resources that can be used in
subsequent function calls.  So $M_f$ corresponds to the stack-frame
size of the function $f$.
%
\begin{center}
${ \{ (Z = \log_2(h-l)) + Z{\cdot}M_\code{bsearch} \} }$
\code{bsearch(x,l,h)}
${ \{ Z{\cdot}M_\code{bsearch} \} }$

\end{center}
%
In the pre and post condition, the logical variable $Z$ is used to
relate the size of the input with the potential that is left after the
function call.  The logical part $Z = \log_2(h-l)$ of the precondition
is $0$ if $Z = \log_2(\state_0(h)-\state_0(l))$ in the initial state
$\state_0$ and $\infty$ otherwise.

Assume for the remainder of this section that we are interested in
bounding the number of \emph{ticks} in a program.  In other words, we
use a resource metric that assigns cost $n$ to the function call
\code{tick(n)} and cost $0$ to all other operations.  The argument $n$
is an integer and a negative number means that $-n$ resources are
returned.

We say that the analysis is amortized because the use of a potential
function $\Phi$ enables us to amortize the cost of operations like in
typical textbook examples that use the potential method of Tarjan's
amortized analysis.  As an illustrative example, \pref{fig:xmplinc}
shows a program that repeatably increments a binary counter that is
implemented with a boolean-valued ($0$ or $1$) array.  The derived
bound in the precondition states that the number of ticks that are
executed by the program are bounded by $\#_1(a) + 2k$.  Here, $k$ is
the number of increments to the counter that are performed by the
program and $ \#_1(a) = \# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ is
the number of $1$'s in the array $a$.  We have inserted the statement
\code{tick(1)} into the program whenever we update the array $a$.

@@

See Section XXX for more a formal definition of the logic and
additional example derivation.

@@

To this end, we fix the shape of the potential functions $\Phi$ so
that it becomes possible to use \emph{linear programming} (LP) to
compute a derivation in the logic.  If we assume that a program state
$\state$ simply maps variable to integers then we require
$$
\Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}} q_{(x,y)} {\cdot} |[\state(x),\state(y)]|
$$
where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  Note that
$q_0$ is the constant potential and that we treat constants as program
variables.  For instance, we always have $q_{(0,x)}|[0,x]|$ as a
component of $\Phi(\state)$ if $x \in \dom{\state}$.
%
We then develop an inference rule for each syntactic construct that
derives sound triples $\htriple{\Phi}P{\Phi'}$ in the quantitative
logic but enables inference by using an LP solver.

This idea is best explained by example.  If we again use the
\emph{ticks} metric that assigns cost $n$ to the function call
\code{tick(n)} and cost $0$ to all other operations then the cost of
the following example can be bounded by $|[x,y]|$.
\begin{lstlisting}
while (x<y) {x=x+1; tick(1);} #\hfill \textnormal{(Example 1)}#
\end{lstlisting}
To derive this bound in our automatic amortized analysis, we start
with the initial potential $\Phi_0 = |[x,y]|$ (short for
$\Phi_0(\state) = |[\state(x),\state(y)]|$) which we also use as the
loop invariant.  For the loop body we then (like in Hoare logic) have
to derive a triple like
$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  However, we can
only do so if we utilize the fact that $x<y$ at the beginning of the
loop body.  The reasoning then works as follows.  We start with the
potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
assignment.  If we denote the updated version of $x$ after the
assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
the potential before and after the assignment \code{x=x+1} holds.
This means that we have the potential $|[x,y]| + 1$ before the
statement \code{tick(1)}.  Since \code{tick(1)} consumes one resource unit,
we end up with potential $|[x,y]|$ after the loop body and have
established the loop invariant again.

\begin{figure}[t]
  \centering
\begin{lstlisting}[mathescape]
${\color{blue}%
  \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
while (x+K<=y) {
  ${\color{blue}%
  x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
  x=x+K; 
  ${\color{blue}%
  \dashv x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
  tick(T);
  ${\color{blue}%
  \dashv x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
}
${\color{blue}%
 \dashv x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
\end{lstlisting}
  \caption{Automatic derivation of a tight bound on the number of
    ticks for a standard \emph{for loop}.  The parameters $K>0$ and
    $T>0$ are not program variables but denote a concrete constants in
    the program.  The reasoning works uniformly for all such $K$ and
    $T$.}
  \label{fig:ex1}
\end{figure}
More formally, we develop (in section XXX) a judgement 
$$
\Gamma; Q \vdash S \dashv Q'; \Gamma'
$$
such that $\Gamma$ contains logical assertions like $x<y$ that we
collect along the conditional branches of the program and $Q$ is a
family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
indexed by the variables that are currently in scope.  The logical
context $\Gamma$ is simply a conjunction of inequalities between
linear combinations variables.  It is used to determine at the
assignment if we can extract constant potential to pay for future
cost.  It is important to note that the reasoning about assertions in
$\Gamma$ are very basic.  We do not perform any fixpoint computations
and only derive trivial loop invariants about variables that are
unchanged in the loop body.

Note that our notion of a potential function is a refinement of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}
  while (x<y) {tick(-1); x=x+1; tick(1);} #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
tick annotations can be seamlessly combined in a larger program.
\begin{lstlisting}
  while (x<y) {x=x+1; tick(10);} #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
%
\pref{fig:ex1} shows a derivation of the bound
$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
version of Example~1 in which increment $x$ be a constant $K>0$ and
consume $T>0$ resources in each iteration.  The reasoning is similar
to the description of Example~1 except that we obtain the potential
$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
assertions in $\Gamma$ are only used in rule for the assignment
\code{x=x+K}.  To the best of our knowledge, no other tool is
currently capable of deriving a tight bound on the cost of such a
loop.  For $T=1$ (many systems focus on the number of loop iterations
without a cost model) and $K=10$, KoAT~\cite{BrockschmidtEFFG14}
computes the bound $x + y + 10$, PUBS~\cite{AlbertAGPZ12} computes
the tight bound $0.1(y-x)$, Rank~\cite{AliasDFG10} computes the bound XXX, and
LOOPUS~\cite{SinnZV14} computes the bound $y-(x+9)$.

To understand the capabilities of the automatic analysis, it is
instructive to study more involved examples.  \pref{fig:ex_list}
contains Examples \emph{speed\_1} and \emph{speed\_2} that are taken
from previous work on bound analysis~\cite{GulwaniMC09}.  The SPEED
tool~\cite{GulwaniMC09} derives the bounds as our analysis but
requires inference of a \emph{disjunctive invariant}.  In the abstract
interpretation community, these invariants are known to be notoriously
difficult to handle.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is more tricky and trying to understand how it
works may be challenging.  However, with the amortized analysis in
mind, using the potential transfer reasoning, it is almost trivial to
prove a bound.  While the SPEED tool has to find a fairly
involved invariant for the loop, our tool is simply reasoning locally
and works without any clever tricks. We obtain the tight bound $|[x,
n]| + |[z, n]|$.

Example \emph{t08a} shows the ability of the analysis to handle
negative and non-negative numbers.  Note that there are no
restrictions on the signs of $z$ and $y$.  We also see that we can
accurately track the size change of $y$ in the first loop.
Furthermore, \emph{t08a} shows that we do not handle the constants $1$
or $0$ in any special way.  In all examples you could replace $0$ and
$1$ with other constants like we did in the second loop and still
derive a tight bound.  The only information, that the analyzer needs
is $y \geq c$ before assigning $y = y - c$.

Example \emph{t27} shows how amortization can be used to handle tricky
nested loops.  The outer loop we increment the variable $n$ till $n =
0$.  In each of the $|[n,0]|$ iterations, we increment the variable
$y$ by $1000$.  We then non-deterministically execute an inner loop
that decrements $y$ by $100$ until $y<100$.  The analysis can derive
that only the first execution of the inner loop depends on the initial
value of $y$.  We again derive tight constant factors.

As mentioned, the analysis naturally handles advanced control flow
like \code{break} and \code{return} statements, and mutual recursion.
\pref{fig:ex_rec} contains two mutually-recursive functions with their
automatically derived tick bounds.  The function \code{count\_down}
decrements its first argument $x$ until it reaches the second argument
$y$. It then recursively calls the function \code{count\_up}.  The
function \code{count\_up} is dual to \code{count\_down}.  Here, we
count up $y$ by $2$ and recursively call \code{count\_down}.  Our
analysis system can compute a tight bound.  On this example, the
analysis amounts to computing the meeting point of two trains that
approach each other with different speeds.
% This is comparable to the
% determination of the point at which to trains meet if they approach
% each other with different speeds.

We only show a very small selections of the programs that we can
handle automatically here.  In \pref{app:cat} you find a list of more
than $30$ classes of challenging programs that we can automatically
analyze.  We hope that this gives you an impression of the
capabilities of the analysis but we also point out that this list is
far from being exhaustive.

\newlength{\progwidth}

@@

  \caption{Derivations of bounds on the number of ticks for
    challenging examples.  In the potential functions, we only mention
    the non-zero terms and in the logical context $\Gamma$ we only
    mention assertions that we actually use in the reasoning. In
    Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]|
    {+} 0.05|[0,y]|$.}
  \label{fig:ex_list}
\end{figure*}

@@

\section{Quantitative Hoare Logic}

@@

\section{Automatic Amortized Analysis}

@@

\section{Experimental Evaluation}

>>>>

\section{Overview and Examples}
\label{sec:inform}

In this section, we informally introduce the quantitative program
logic and the automatic amortized analysis for for Clight programs.
The purpose of this section is to build some intuition that should
make the following sections easy to read.

@@

The idea that drives the design of our framework is amortized
analysis~\cite{Tarjan-amort}.  Assume that a program $S$ executes on a
starting state $\state$ and consumes $n$ resource units of some
quantity we are interested in.  We denote that by writing $(S, \state)
\Downarrow_n \state'$ where $\state'$ the program state after the
execution.  The basic idea of amortized analysis is to define a
\emph{potential function} $\Phi$ that maps program states to non-negative
numbers and to show that $\Phi(\state) \geq n$ if $\state$ is a
program state such that such that $(S, \state) \Downarrow_n \state'$.

To obtain a compositional analysis we also have to take into account
the state resulting from the program execution.  We thus use two
potential functions, one that applies before the execution, and one
that applies after.  The two functions must respect the relation
$\Phi(\state) \ge n + \Phi'(\state')$ for all states $\state$ and
$\state'$ such that $(S, \state) \Downarrow_n \state'$.  Intuitively,
$\Phi(\state)$ must provide enough \emph{potential} for both, paying
for resource cost of the computation and paying for the potential
$\Phi'(\state')$ on the resulting state $\state'$. That way, if
$(\state, S_1) \Downarrow_n \state'$ and $(\state', S_2) \Downarrow_m
\state''$, we get $\Phi(\state) \ge n + \Phi'(\state')$ and
$\Phi'(\state') \ge m + \Phi''(\state'')$.  This can be composed to
$\Phi(\state) \ge (n + m) + \Phi''(\state'')$.  Note that the initial
potential function $\Phi$ provides an upper bound on the resource
consumption of the whole program.  What we have observed is that, if
we define $\htriple{\Phi} {S}{\Phi'}$ to mean
$$
\forall \state\, n\, \state'.\, (\state, p) \Downarrow_n \state' \implies \Phi(\state) \ge
n + \Phi'(\state')
$$
then we get the following familiar looking rule.
$$
\RuleNolabel
{\htriple{\Phi}{S_1}{\Phi'} \\ \htriple{\Phi'}{S_2}{\Phi''}}
{\htriple{\Phi} {S_1; S_2}  {\Phi''}}
$$
%
Similarly, other language constructs lead to rules for the potential
functions that look very similar to Hoare logic or effect system
rules.  These rules enable interactive reasoning about resource usage
in a flexible and compositional way, which, as a side effect, produces
a certificate for the derived resource bound.

% We have implemented and proved sound such a quantitative Hoare for
% CompCert Clight in the Coq Proof Assistant.  Because we use a shallow
% embedding in Coq, every function that can be expressed in Coq can in
% principle be used as a resource bound.

It is also possible to incorporate boolean preconditions into the
bounds to express that the bound is only valid for a certain class of
inputs.  To this end, we allow the potential function $\Phi$ in the
pre and post conditions to take non-negative number or $\infty$
(infinity) as values.  Infinity, plays the same role as
$\mathit{false}$ in Hoare logic.  Boolean assertions can be embedded
into potential functions by mapping $\mathit{false}$ to $\infty$ and
$\mathit{true}$ to $0$.

@@

\caption{Example derivation using amortized reasoning.  We write
  $\#_1(a)$ for $\# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ and use the
  tick metric that assigns cost $n$ to the statement \lstinline{tick(n)} and
  0 to all other operations.
  }
\label{fig:xmplinc}
\end{figure}

We can use the logic to prove a concrete bound for a given resource
metric, or a bound that is parametric in a set of resource metrics.
For example, it is possible to derive the following quantitative
triple for a binary-search function \code{bsearch} that holds for all
stack metrics $M$.  
%
\begin{center}
${ \{ (Z = \log_2(h-l)) + Z{\cdot}M_\code{bsearch} \} }$
\code{bsearch(x,l,h)}
${ \{ Z{\cdot}M_\code{bsearch} \} }$
\end{center}
%
A stack metric assigns cost $0$ to all evaluation steps except
function calls.  Before a call $f(x)$, we consume $M_f > 0$ resources
and after the call we return $M_f$ resources that can be used in
subsequent function calls.  So $M_f$ corresponds to the stack-frame
size of the function $f$.  In the pre- and postcondition, the logical
variable $Z$ is used to relate the size of the input with the
potential that is left after the function call.  The logical part $Z =
\log_2(h-l)$ of the precondition is $0$ if $Z =
\log_2(\state_0(h)-\state_0(l))$ in the initial state $\state_0$ and
$\infty$ otherwise.

For simplicity, assume for the remainder of this section that we are
interested in bounding the number of \emph{ticks} in a program.  In
other words, we use a resource metric that assigns cost $n$ to the
function call \code{tick(n)} and cost $0$ to all other operations.
The argument $n$ is an integer and a negative number means that $-n$
resources are returned.

We say that the analysis is amortized because the use of a potential
function $\Phi$ enables us to amortize the cost of operations like in
typical textbook examples that use the potential method of Tarjan's
amortized analysis.  As an illustrative example, \pref{fig:xmplinc}
shows a program that repeatably increments a binary counter that is
implemented with a boolean-valued ($0$ or $1$) array.  The derived
bound in the precondition states that the number of ticks that are
executed by the program are bounded by $\#_1(a) + 2k$.  Here, $k$ is
the number of increments to the counter that are performed by the
program and $ \#_1(a) = \# \{ i \mid 0 \le i < N \land a[i] = 1 \}$ is
the number of $1$'s in the array $a$.  We have inserted the statement
\code{tick(1)} into the program whenever we update the array $a$.  A
naive analysis of the algorithm would lead to a quadratic bound since
there can be a linear number of updates in the inner loop.

@@

See \pref{sec:logic} for more a formal definition of the logic and
additional example derivation.

@@

To this end, we fix the shape of the potential functions $\Phi$ so
that it becomes possible to use \emph{linear programming} (LP) to
compute a derivation in the logic.  If we assume for now that a
program state $\state$ simply maps variable to integers then we
require
$$
\Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}} q_{(x,y)} {\cdot} |[\state(x),\state(y)]|
$$
where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  Note that
$q_0$ is the constant potential and that we treat constants as program
variables.  For instance, we always have $q_{(0,x)}|[0,x]|$ as a
component of $\Phi(\state)$ if $x \in \dom{\state}$.
%
We then develop an inference rule for each syntactic construct that
derives sound triples $\htriple{\Phi}P{\Phi'}$ in the quantitative
logic but enables inference by using an LP solver.

This idea is best explained by example.  If we again use the
\emph{ticks} metric that assigns cost $n$ to the function call
\code{tick(n)} and cost $0$ to all other operations then the cost of
the following example can be bounded by $|[x,y]|$.
\begin{lstlisting}
while (x<y) {x=x+1; tick(1);} #\hfill \textnormal{(Example 1)}#
\end{lstlisting}
To derive this bound in our automatic amortized analysis, we start
with the initial potential $\Phi_0 = |[x,y]|$ (short for
$\Phi_0(\state) = |[\state(x),\state(y)]|$) which we also use as the
loop invariant.  For the loop body we then (like in Hoare logic) have
to derive a triple like
$\htriple{\Phi_0}{\code{x=x+1;tick(1)}}{\Phi_0}$.  However, we can
only do so if we utilize the fact that $x<y$ at the beginning of the
loop body.  % The reasoning then works as follows. We start with the
% potential $|[x,y]|$ and the fact that $|[x,y]| > 0$ before the
% assignment.  
If we denote the updated version of $x$ after the
assignment by $x'$ then the relation $|[x,y]| = |[x',y]| + 1$ between
the potential before and after the assignment \code{x=x+1} holds.
This means that we have the potential $|[x,y]| + 1$ before the
statement \code{tick(1)}.  Since \code{tick(1)} consumes one resource unit,
we end up with potential $|[x,y]|$ after the loop body and have
established the loop invariant again.

Note that our notion of a potential function is a generalization of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}
  while (x<y) {tick(-1); x=x+1; tick(1);} #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
tick annotations can be seamlessly combined in a larger program.
\begin{lstlisting}
  while (x<y) {x=x+1; tick(10);} #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
%
Also note that there are no restrictions on the signs of the input
variables in the examples.

\begin{figure}[t]
  \centering
\begin{lstlisting}[mathescape]
${\color{blue}%
  \cdot; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
while (x+K<=y) {
  ${\color{blue}%
  x + K \le y; \; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
  x=x+K; 
  ${\color{blue}%
  \dashv x \leq y; \, T + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| \vdash }$
  tick(T);
  ${\color{blue}%
  \dashv x \leq y; \, 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]| }$
}
${\color{blue}%
 \dashv x \geq y; 0 + \frac{T}{K} {\cdot} |[x,y]| + 0 {\cdot} |[y,x]|}$
\end{lstlisting}
  \caption{Automatic derivation of a tight bound on the number of
    ticks for a standard \emph{for loop}.  The parameters $K>0$ and
    $T>0$ are not program variables but denote a concrete constants in
    the program.  The reasoning works uniformly for all such $K$ and
    $T$.}
  \label{fig:ex1}
\end{figure}
%
\newlength{\progwidth}

@@

  \caption{Derivations of bounds on the number of ticks for
    challenging examples.  In the potential functions, we only mention
    the non-zero terms and in the logical context $\Gamma$ we only
    mention assertions that we actually use in the reasoning. In
    Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]|
    {+} 0.05|[0,y]|$.}
  \label{fig:ex_list}
\end{figure*}
%
%
More formally, we develop (\pref{sec:AAA}) a judgement 
$$
\Gamma; Q \vdash S \dashv Q'; \Gamma'
$$
such that $\Gamma$ contains logical assertions like $x<y$ that we
collect along the conditional branches of the program and $Q$ is a
family of coefficents $Q = (q_{(a,b)})_{a,b \in \text{scope}}$ that is
indexed by the variables that are currently in scope.  The logical
context $\Gamma$ is simply a conjunction of inequalities between
linear combinations variables.  It is used to determine at the
assignment if we can extract constant potential to pay for future
cost.  It is important to note that the reasoning about assertions in
$\Gamma$ is very basic.  We do not perform any fixpoint computations
and only derive trivial loop invariants about variables that are
unchanged in the loop body.

\pref{fig:ex1} shows a derivation of the bound
$\frac{T}{K}{\cdot}|[x,y]|$ on the number of ticks for a generalized
version of Example~1 in which increment $x$ be a constant $K>0$ and
consume $T>0$ resources in each iteration.  The reasoning is similar
to the description of Example~1 except that we obtain the potential
$K{\cdot}\frac{T}{K}$ after the assignment.  Note that the logical
assertions in $\Gamma$ are only used in rule for the assignment
\code{x=x+K}.  To the best of our knowledge, no other tool for C is
currently capable of deriving a tight bound on the cost of such a
loop.  For $T=1$ (many systems focus on the number of loop iterations
without a cost model) and $K=10$, KoAT~\cite{BrockschmidtEFFG14}
computes the bound $x + y + 10$, Rank~\cite{AliasDFG10} computes the
bound $y-x-7$, and LOOPUS~\cite{SinnZV14} computes the bound
$y-(x+9)$.  Only PUBS~\cite{AlbertAGPZ12} computes the tight bound
$0.1(y-x)$ if we translate the program into a term-rewriting system by
hand.

As mentioned, the automatic analysis can handle challenging example
programs without special tricks or techniques.  Examples
\emph{speed\_1} and \emph{speed\_2} that are taken from previous work
on bound analysis~\cite{GulwaniMC09} and demonstrate the our method
can handle \emph{tricky iteration patterns}.  The SPEED
tool~\cite{GulwaniMC09} derives the same bounds as our analysis but
requires inference of \emph{disjunctive invariants}.  In the abstract
interpretation community, these invariants are known to be notoriously
difficult to handle.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is more tricky and we found it hard to
manually find a bound.  However, with the amortized analysis in mind,
using the potential transfer reasoning, it is almost trivial to prove
the tight bound $|[x, n]| + |[z, n]|$.

Example \emph{t08a} shows the ability of the analysis to discover
interaction between \emph{sequential loops or function calls} through
size changes of variables.  We accurately track the size change of $y$
in the first loop by transferring the potential $0.1$ form $|[y,z]| to
|[0,y]|$.  Furthermore, \emph{t08a} shows that we do not handle the
constants $1$ or $0$ in any special way.  In all examples you could
replace $0$ and $1$ with other constants like we did in the second
loop and still derive a tight bound.  The only information, that the
analyzer needs is $y \geq c$ before assigning $y = y - c$.

Example \emph{t27} shows how amortization can be used to handle
\emph{interacting nested loops}.  The outer loop we increment the variable $n$
till $n = 0$.  In each of the $|[n,0]|$ iterations, we increment the
variable $y$ by $1000$.  We then non-deterministically execute an
inner loop that decrements $y$ by $100$ until $y<100$.  The analysis
can derive that only the first execution of the inner loop depends on
the initial value of $y$.  We again derive tight constant factors.
%

As mentioned, the analysis also handles advanced control flow like
\code{break} and \code{return} statements, and mutual recursion.
\pref{fig:ex_rec} contains two mutually-recursive functions with their
automatically derived tick bounds.  The function \code{count\_down}
decrements its first argument $x$ until it reaches the second argument
$y$. It then recursively calls the function \code{count\_up}.  The
function \code{count\_up} is dual to \code{count\_down}.  Here, we
count up $y$ by $2$ and recursively call \code{count\_down}.  Our
analysis system is the only available system that can compute a tight
bound on this example.  The analysis amounts to computing the meeting
point of two trains that approach each other with different speeds.
% This is comparable to the
% determination of the point at which to trains meet if they approach
% each other with different speeds.

We only show a small selections of the programs that we can handle
automatically here.  In \pref{app:cat} is a list of more than $30$
classes of challenging programs that we can automatically analyze.
\pref{sec:exper} contains a more detailed comparison with other tools
for automatic bound derivation.

@@

\section{Quantitative Hoare Logic}
\label{sec:logic}

@@

\section{Automatic Amortized Analysis}
\label{sec:AAA}

@@

\section{Experimental Evaluation}
\label{sec:exper}
